\bf Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-07-16 manuscript/main.tex
basis notation revised

\subsubsection{Finite-basis approximation of the predictors and regression coefficients}\label{sec: finite basis approximation}	
Given a twice-differentiable basis $\{b_l(t)\}$ of $L^2([0,1])$,
the functional predictors,
the corresponding regression coefficients,
and PLS component direction
 can be decomposed into
	\begin{equation*}
		X_{ij}(t) = \sum_{l=1}^\infty \theta_{ijl} \, b_l(t),~
		\beta_j(t) = \sum_{l=1}^\infty \eta_{0,jl} \, b_l(t), j=1, \ldots, K,~
		%%%%%%%%%%%%%%%%%%%%%%
		\xi 
		=
		\sum_{l=1}^\infty 
		b_l(t)
	\end{equation*}
	At the $l$-th iteration of the PLS algorithm, let
	\begin{equation}
		( \hat{\boldsymbol{\gamma}}_1^\top, \ldots, \hat{\boldsymbol{\gamma}}_K^\top, \hat{\boldsymbol{\xi}}_{K+1}^\top )^\top
	\end{equation}

	where $\theta_{ijl}$ and $\eta_{0,jl}$ are the coefficients of $X_{ij}(t)$ and $\beta_j(t)$ corresponding to the $l$th basis function $b_l(t)$, respectively.
Since functional predictors are observed at finitely many discrete time points in practice,
we use finite-dimensional projections obtained by the expansion truncating at $M$ terms.
We choose a moderately large value of $M$ (e.g., 15 or 20) to capture rich functional variation without requiring fine-tuning, as our method promotes smoothness through penalization (see Section~\ref{sec: Regularization of PLS Components}). We denote the truncated space by
	\begin{equation}
		\mathcal{B} := \operatorname{span} \bigl( b_1(t), \ldots, b_M(t) \bigr).
	\end{equation}
	 We denote the  predictors projected onto $\mathcal{B}$ and the resulting coefficients as
	\begin{equation*}
		\widetilde{X}_{ij}(t) = \sum_{l=1}^M \theta_{ijl} \, b_l(t), \quad 
		\tilde{\beta}_j(t) = \sum_{l=1}^M \eta_{0,jl} \, b_l(t).
	\end{equation*}
	The $i$th  hybrid predictor with projected functional parts, is a tuple, denoted as
	\begin{equation*}
		\widetilde{W}_i := 
		(
		X_{i1}, \ldots, X_{iK}, \boldsymbol{Z}_i
		).
	\end{equation*}
	We adopt the following notation for the coefficients:
	\begin{equation}
		\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top \in \mathbb{R}^M,
		\quad
		\boldsymbol{\eta}_{0j} := (\eta_{0,j1}, \ldots, \eta_{0,jM})^\top  \in \mathbb{R}^M,
	\end{equation}
	and
	\begin{equation}
		\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
	\end{equation}
The observations in the hybrid PLS problem can be compactly represented by the tuple
\begin{equation}\label{def:problem_instance}
	(\mathbb{B}, \mathbb{B}^{\prime \prime}, \Theta, \mathbf{y})
	\in
	\mathbb{R}^{(MK+p) \times (MK+p)}
	\times 
	\mathbb{R}^{(MK+p) \times (MK+p)}
	\times
	\mathbb{R}^{n \times (MK+p)}
	\times
	\mathbb{R}^{n}
	,
\end{equation}
defined as follows.
First, let $B, B^{\prime\prime} \in \mathbb{R}^{M \times M}$ denote the Gram matrices of the basis functions and their second derivatives, respectively, with entries  
\begin{equation}
	B_{m, m'} := \int_0^1 b_m(t)\, b_{m'}(t)\, dt,
	\quad
	B^{\prime\prime}_{m, m'} := \int_0^1 b^{\prime\prime}_m(t)\, b^{\prime\prime}_{m'}(t)\, dt,
\end{equation}
for $m, m' = 1, \ldots, M$.
Leveraging these notations, 
we define the $(MK + p) \times (MK + p)$ block-diagonal matrices
\begin{equation}\label{def:gram_block}
\mathbb{B} := \operatorname{blkdiag}(B, \ldots, B, I_p)
\quad \text{and} \quad
\mathbb{B}^{\prime\prime} := \operatorname{blkdiag}(B^{\prime\prime}, \ldots, B^{\prime\prime}, I_p),
\end{equation}
We denote $\Theta := (\Theta_1, \ldots, \Theta_K, Z)
\in
\mathbb{R}^{n \times (MK+p)}
$ , and let $\mathbf{y} := (y_1, \ldots, y_n)^\top$ denote the observed response vector.

	\begin{remark}
While it is possible to use a different basis for each functional predictor, we adopt a common basis for all predictors to simplify the notation. However, the definition of 
$\mathbb{B}$
and
$\mathbb{B}^{\prime \prime}$
 remains general enough to accommodate different bases across the $K$ predictors.
\end{remark}

>>>>

\subsubsection{Finite-basis approximation of the predictors and regression coefficients}\label{sec: finite basis approximation}	
Given a twice-differentiable basis $\{b_m(t)\}$ of $L^2([0,1])$,  the $j$th functional predictor, regression coefficient, and PLS direction (suppressing iteration indices) are represented as
	\begin{equation*}
		X_{ij}(t) = \sum_{m=1}^\infty \theta_{ijm} \, b_m(t),\quad
		\beta_j(t) = \sum_{m=1}^\infty \eta_{jm} \, b_m(t),\quad
		%%%%%%%%%%%%%%%%%%%%%%
		\xi_j(t)
		=
		\sum_{m=1}^\infty
		\gamma_{jm}
		b_m(t)
	\end{equation*}	
In practice, the full set of coefficients can not be obtained, as functional data are measured on a finite grid. Thus, we truncate the expansion at $M$ terms and define the truncated basis space
\begin{equation}
	\mathcal{B} := \operatorname{span}\bigl(b_1(t), \ldots, b_M(t)\bigr).
\end{equation}
We choose a moderately large $M$ (e.g., 15 or 20) to capture functional variation without fine-tuning, as smoothness is handled via penalization (see Section~\ref{sec: Regularization of PLS Components}).

	The truncated expansions of the predictor and coefficient are denoted as
	\begin{equation*}
		\widetilde{X}_{ij}(t) := \sum_{m=1}^M \theta_{ijm} \, b_m(t), \quad
		\tilde{\beta}_j(t) := \sum_{m=1}^M \eta_{jm} \, b_m(t),
	\end{equation*}
	and our suggest method restricts each PLS direction to admit the following expansion:
	\begin{equation*}
		\xi_j(t) = \sum_{m=1}^M \gamma_{jm} \, b_m(t).
	\end{equation*}
The $i$th hybrid predictor, with its functional parts projected onto the $\mathcal{B}$, is represented by the tuple
	\begin{equation*}
		\widetilde{W}_i := 
		(
		\widetilde{X}_{i1}, \ldots, \widetilde{X}_{iK}, \boldsymbol{Z}_i
		).
	\end{equation*}
Let $\boldsymbol{\theta}_{ij}$, $\boldsymbol{\eta}_j$, and $\boldsymbol{\gamma}_j$ denote the $M$-dimensional vectors of coefficients:
	\begin{equation}
		\boldsymbol{\theta}_{ij} := (\theta_{ij1}, \ldots, \theta_{ijM})^\top,
		\quad
		\boldsymbol{\eta}_{j} := (\eta_{j1}, \ldots, \eta_{jM})^\top ,
			\boldsymbol{\gamma}_{j} := (\gamma_{j1}, \ldots, \gamma_{jM})^\top.
	\end{equation}
For the predictors, we stack the coefficient vectors across observations into the matrix
	\begin{equation}
		\Theta_j := (\boldsymbol{\theta}_{1j}, \ldots, \boldsymbol{\theta}_{nj})^\top  \in \mathbb{R}^{n \times M}.
	\end{equation}
The full data for the hybrid PLS problem can be represented by the tuple
\begin{equation}\label{def:problem_instance}
	(\mathbb{B}, \mathbb{B}^{\prime\prime}, \Theta, \mathbf{y}) \in 
	\mathbb{R}^{(MK+p) \times (MK+p)} \times 
	\mathbb{R}^{(MK+p) \times (MK+p)} \times 
	\mathbb{R}^{n \times (MK+p)} \times 
	\mathbb{R}^n,
\end{equation}
where $B, B^{\prime\prime} \in \mathbb{R}^{M \times M}$ are the Gram matrices of the basis functions and their second derivatives, with entries
\begin{equation}
	B_{m, m'} := \int_0^1 b_m(t)\, b_{m'}(t)\, dt, \quad
	B^{\prime\prime}_{m, m'} := \int_0^1 b_m''(t)\, b_{m'}''(t)\, dt,
\end{equation}
for $m, m' = 1, \ldots, M$. We then define the block-diagonal matrices
\begin{equation}\label{def:gram_block}
	\mathbb{B} := \operatorname{blkdiag}(B, \ldots, B, I_p), \quad
	\mathbb{B}^{\prime\prime} := \operatorname{blkdiag}(B^{\prime\prime}, \ldots, B^{\prime\prime}, I_p),
\end{equation}
the full design matrix $\Theta := (\Theta_1, \ldots, \Theta_K, Z) \in \mathbb{R}^{n \times (MK + p)}$, and the response vector $\mathbf{y} := (y_1, \ldots, y_n)^\top$.

\begin{remark}
	While different bases could be used for each functional predictor, we adopt a common basis for simplicity. The definitions of $\mathbb{B}$ and $\mathbb{B}^{\prime\prime}$ remain general enough to accommodate distinct bases if needed.
\end{remark}
